Stability and Change in Personal Culture

### 2020-02-12 manuscript9.Rmd
final edits

On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this does not need to be the case. The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse], and suggest that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse]. As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.

@@

Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.

@@

Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list. In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.

@@

We will discuss in more detail below how different items perform. To answer our first question, however, we need only focus on the overall distribution. Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1, meaning that for almost all items, measurement error or non-persistent change tends to be much more common than persistent change. We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5. This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.

@@

Our second research question asks whether there is evidence that younger respondents are updating their views more than older respondents. Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.

Of the 108 items that showed evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not. Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.

@@

There is a lot to process, even in this summary figure. The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low in absolute terms. Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity. In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]

@@

While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently. Views on abortion, especially when considered as a scale, are consistently reported. As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change. As discussed previously, views on gay marriage display high levels of active updating. In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating. This is notable, as public opinion and policies regarding both have shifted considerably in recent years. Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell]. Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana use have been driven primarily by cohort replacement.

@@

In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession. Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood. The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time. In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating. Even those items that did display strong evidence for active updating (e.g. "whites rich or poor" and "whites work hard") were still hard to predict wave after wave.

@@

Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults. Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production" that give rise to persistent beliefs in adulthood.

>>>>

On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this need not be the case. The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse]. Such accounts imply that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse]. As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.

@@

Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as publicly joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.

@@

Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list. In other words, even for items that show strong evidence of some active updating, the overall amount of attitude change in the population is likely small.

@@

We will discuss in more detail below how different items perform. To answer our first question, however, we need only focus on the overall distribution. Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1. This means that for almost all items, measurement error or non-persistent change tend to be much more common than persistent change. We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5. This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.

@@

Our second research question asks whether there is evidence that younger respondents update their views more than older respondents. Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.

Of the 108 items that showed any evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not. Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.

@@

There is a lot to process, even in this summary figure. The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low. Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity. In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]

@@

While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently. Views on abortion, especially when considered as a scale, are consistently reported. As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change. As discussed previously, views on gay marriage display high levels of active updating. In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating. This is notable, as public opinion and policies regarding both have shifted considerably in recent years. Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell]. Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.

@@

In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession. Most beliefs about gender roles, sexual morality, and abortion appear to be settled by early adulthood. The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time. In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating. Even those items that did display strong evidence for active updating (e.g. "whites rich or poor" and "whites work hard") were still hard to predict wave after wave.

@@

Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults. Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production"—childhood and adolescence—that give rise to persistent beliefs in adulthood.
